---
title: "Normalization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Normalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
'%>%' <-  kableExtra::'%>%'

```
```{r load library, echo=F, eval=T, warning=F, message=F}
library(AnalyzAIRR)
```


# Down-sampling

This strategy can be applied when the studied samples largely differ in their repertoire sizes.

By using the `sampleRepSeqExp()` function, users can choose the value to which all the samples are downsampled. If not specified, the lowest number of sequences across the dataset will be used.

This function returns a new RepSeqExperiment object with the downsized data.

```{r downsample, echo=T}
RepSeqData_ds<- sampleRepSeqExp(x = RepSeqData, 
                                sample.size = 50000)
```

```{r, echo=F, fig.align='center', fig.width=10, fig.height=6}
margin_r = ggplot2::theme(plot.margin = ggplot2::unit(c(0,1,0,0), "cm"))
margin_l = ggplot2::theme(plot.margin = ggplot2::unit(c(0,0,0,1), "cm"))

p3<- plotStatistics(x = RepSeqData_ds, stat = "nSequences", colorBy = "sample_id")+margin_l
p4<- plotStatistics(x = RepSeqData_ds, stat = "aaClone", colorBy = "sample_id")+margin_l
p1<- plotStatistics(x = RepSeqData, stat = "nSequences", colorBy = "sample_id")+margin_r
p2<- plotStatistics(x = RepSeqData, stat = "aaClone", colorBy = "sample_id")+margin_r

gridExtra::grid.arrange(gridExtra::arrangeGrob(p1,p2, top = grid::textGrob("Pre-downsampling", x = .55)), gridExtra::arrangeGrob(p3,p4, top = grid::textGrob("Post-downsampling", x = .62)),ncol=2)

```
<figcaption align = "center"><em class="special">Figure 1: Summary plots showing the effect of downsampling curves plotting the number of sequences and aaClones in each sample</em></figcaption>
<br>

# Shannon-based normalization

This strategy adapted from [Chaara et al., 2018](https://doi.org/10.3389/fimmu.2018.01038) can be used to eliminate “uninformative” sequences resulting from experimental noise. It uses the Shannon entropy as a threshold and is applied at the `ntClone` level. 
This strategy is particularly efficient when applied on small samples as it corrects altered count distributions caused by a high-sequencing depth.

The function `ShannonNorm()` allows the application of this strategy without the need to specify any parameter and returns a new RepSeqExperiment object with the corrected data.

```{r shannon norma, echo=T, eval=T}
RepSeqData_sh <- ShannonNorm(x = RepSeqData)
```

```{r, echo=F,fig.align='center', fig.width=10, fig.height=3}
RepSeqData_sh@assayData$sample_id<- paste0(RepSeqData_sh@assayData$sample_id,"_norm")
RepSeqData_sh@metaData$sample<- RepSeqData_sh@metaData$sample_id
RepSeqData_sh@metaData$sample_id<- as.factor(paste0(RepSeqData_sh@metaData$sample_id,"_norm"))
RepSeqData_sh@metaData$normalized<- "normalized"
RepSeqData_sh@metaData$normalized<- as.factor(RepSeqData_sh@metaData$normalized)
rownames(RepSeqData_sh@metaData)<- RepSeqData_sh@metaData$sample_id

RepSeqData@metaData <- RepSeqData@metaData %>% dplyr::select(-chao1 ,-iChao)
RepSeqData@metaData$sample<- RepSeqData@metaData$sample_id
RepSeqData@metaData$normalized<- "original"
RepSeqData@metaData$normalized<- as.factor(RepSeqData@metaData$normalized)

norm<- mergeRepSeq(RepSeqData,RepSeqData_sh)

plotStatistics(x = norm, stat = "ntClone",  colorBy = "normalized", facetBy = "sample")+
  ggplot2::theme(axis.text.x = ggplot2::element_blank())

# p1<-plotStatistics(x = norm, stat = "nSequences",  colorBy = "normalized", facetBy = "sample")+
#   ggplot2::theme(axis.text.x = ggplot2::element_blank())

# 
# p2<- plotDiversity(x = norm, index="shannon", level = "aaClone",colorBy = "normalized",  facetBy = "sample")+
#   ggplot2::theme(axis.text.x = ggplot2::element_blank())
# 
# 
# gridExtra::grid.arrange(p1,p2, ncol=1)


```
<figcaption align = "center"><em class="special">Figure 2: A summary plot showing the number of ntClones pre-and post-Shannon normalizatin in each sample</em></figcaption>
<br>

**Notes**: 

  + Eliminated sequences in each sample are stored in the **otherData** slot.
  + Chao, the richness estimator, isn't recalculated for normalized datasets as as their original composition has been modified.

