---
title: "Data manipulation"
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
'%>%' <-  kableExtra::'%>%'

```
```{r load library, echo=F, eval=T, warning=F, message=F}
library(AnalyzAIRR)
```


Different functions were developed to filter out or extract samples or sequences within a RepSeqExperiment object.


# Filtering

## Sequence filtering

### Based on sequence occurrence

  `filterCount()` filters out sequences, at any level, having a count below a chosen threshold. The function returns a RepSeqExperiment object that can be used to perform in-depth analyses on the remaining dataset.
  
```{r, eval=FALSE}
RepSeqData_filtered <- filterCount(x = RepSeqData,
                                   n = 1,
                                   level = "aaClone")
```

The parameter `group` allows the selection of a group of samples on which the filtering must be applied.
These samples should all belong to a particular group in the **metaData** slot. In this the following case, we're filtering out aaClones with a count of 1 in all amTreg samples.

```{r, eval=FALSE}
RepSeqData_filtered_amTreg <- filterCount(x = RepSeqData,
                                          n = 1,
                                          level = "aaClone",
                                          group = c("cell_subset", "amTreg"))
```


### Based on sequence name

  `filterSequence()` filters out specific sequences in all or a in group of samples.

```{r, eval=FALSE}
RepSeqData_filtered <- filterSequence(x = RepSeqData,
                                      level = "aaClone",
                                      name = "TRAV11 CVVGDRGSALGRLHF TRAJ18",
                                      group = c("cell_subset" , "Teff"))
```


## Sample filtering

  `dropSamples()` offers the possibility to filter out one or multiple repertoires by specifying their corresponding sample_id. 
 These repertoires can be ones identified, for instance, as outliers during the exploratory analysis 

```{r, eval=FALSE}
RepSeqData_drop <- dropSamples(x = RepSeqData,
                              sampleNames=c("tripod-30-813", "tripod-30-815"))

```


# Data selection

### Based on sequence sharing

`getPublic()` allows to subset a RepSeqExperiment object in order to extract sequences that are either shared by:

  + samples belonging to a specified group
  + samples within the whole dataset if the `group` parameter is not specified.
  <br>

The sharing threshold is set to 50% of the selected samples.

```{r getoverlaps,  echo = T, eval=F}
# Get clones present in at least 50% of the samples belonging to the amTreg group
RepSeqData_public <- getPublic(x = RepSeqData, 
                                level = "aaClone", 
                                group = c("cell_subset", "amTreg"))
```

  Similarly, it is possible to extract private sequences with `getPrivate()`. If the parameter `singletons` is set to TRUE, only private sequences with a count of 1 will be returned.
  
```{r, eval=FALSE}
RepSeqData_private <- getPrivate(x = RepSeqData,
                                level = "ntClone",
                                singletons = FALSE)
```


### Based on sequence occurrence
  
  `getTopSequences()` allows the extraction of the top most expressed sequences.
   The `prop` parameter allows users to specify the percentage of top sequences to extract. 
   
```{r, eval=FALSE}
RepSeqData_top <- getTopSequences(x = RepSeqData,
                                  level = "aaClone",
                                  group = c("cell_subset", "Teff"), 
                                  prop = 0.1)

```
   

### Based on sequence functionality

  `getProductive()` and `getUnproductive()` allow the extraction of productive or unproductive sequences respectively in case no filters were applied during the building of the RepSeqExperiment object.


**Note**: All the above-mentioned functions can be applied at any of the following repertoire level: **clone**, **clonotype**, **CDR3aa** and **CDR3nt**.


 