---
title: "Data loading"
---

```{r, include = FALSE, eval=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
'%>%' <-  kableExtra::'%>%'

```

```{r load library, echo=F, eval=T, warning=F, message=F}
library(AnalyzAIRR)
```

## Input format

**AnalyzAIRR** proposes two different data loading processes depending on the alignment files format. 

### Automatic format detection

  **AnalyzAIRR** integrates parser functions to automatically detect the following formats:
  
  * **MiXCR**: https://github.com/milaboratory/mixcr

  * **immunoSEQ**: https://www.immunoseq.com/

  * **MiAIRR**: https://docs.airr-community.org/en/latest/datarep/rearrangements.html 
  
These alignment files can be loaded using the `readAIRRSet()` function. 

### Reformatting of annotated files

**AnalyzAIRR** offers the possibility to load alignment files in other formats than the ones cited in the previous section. These files should contain the following required column names:

- **sample_id**: Sample names

- **V**: Variable gene name (IMGT nomenclature)

- **J**: Joining gene name (IMGT nomenclature)

- **ntCDR3**: Nucleotide CDR3 sequence

- **aaCDR3**: Amino acid CDR3 sequence

- **count**: The occurrence of each sequence within the sample

**Note**: The order of the columns must be respected. Input files can contain additional columns that will not be taken into account during the loading process.

`readFormatSet()` is the equivalent function of `readAIRRSet()` when using user-formatted files. It allows the loading and filtering of a list of files using the same previously described parameters.

## Data loading
### Paths to input files

Specify the list of paths to the aligned files.

A list of 8 samples were selected from a published study (Mhanna et al., 2021) to illustrate the use of this package. Samples were collected from the spleen of 4 healthy mice and sorted into naïve regulatory T cells (nTreg), and activated/memory regulatory T cells (amTreg). 
Fastq files were aligned using MiXCR and exported in a MiAIRR format, and can be found at https://github.com/i3-unit/AnalyzAIRR/tree/master/inst/extdata/MiAIRR

```{r load aligned files, echo=T, eval=T}
l <- list.files(system.file(file.path('extdata/MiAIRR'),
                            package = 'AnalyzAIRR'),
                            full.names = T)
```

### Metadata

It is possible to provide a metadata if users wish to perform inter-group comparative analyses. 

The metadata should be provided as a dataframe containing:

  * a column with the sample names that match the name of the alignment files and their order in the list.

  Only one column containing the sample names should be provided in the metadata file. This column is assigned to the `row.names` argument when loading the metadata. 

  * any additional columns with relevant information for the analyses. Columns could encompass the experimental conditions, clinical variables, etc... 
  
  No specific column names nor order are required.
  
  **Note**: Since **R 4.0.0**, `stringsAsFactors` is set to **FALSE** by default, user will thus need to coerce character vectors into factors.


```{r load metadata}
metadata <- read.table(system.file(file.path('extdata/sampledata.txt'),
                            package = 'AnalyzAIRR'), 
                            header = T,
                            row.names = 1)
metadata$cell_subset <- factor(metadata$cell_subset)
metadata$sex <- factor(metadata$sex)
```

```{r view metadata, echo=F, fig.align='center'}
reactable::reactable(metadata,  bordered = TRUE,  fullWidth = TRUE,
          defaultColDef = reactable::colDef(vAlign = "center",  align = "center", 
                                 headerStyle = list(background = "#f7f7f8")))
```

### Loading

Whether you're using the `readAIRRSet()` or `readFormatSet()` function, the following parameters must be specified:

 * `fileList`: The list of paths to the input files. File extensions can be one of the following: .tsv, txt.gz, .zip, .tar
 
 * `fileFormat`: The name of the tool that was used to generate the alignment files. Should be one of "MiXCR", "immunoseq" or "MiAIRR".
 
 * `chain`: The chain type to be analyzed. Should be one of the following:
   - "TRA", "TRB","TRG" or "TRD" for the TCR repertoires
   - "IGH","IGK" or "IGL" for the BCR repertoires
   
  Only a single chain can be loaded at once. 
 
 
 * `sampleinfo`: A data frame with the metadata information
 
 * `keep.ambiguous`: A choice whether ambiguous sequences should be left in the datasets
 
 * `keep.unproductive`: A choice whether unproductive sequences should be left in the datasets
 
 * `filter.singletons`: A choice whether sequences with an occurrence of 1 should be filtered out
  
 * `aa.th`: A threshold determining the amino acid CDR3 sequence length limits. Refer to the function's documentation for more details
 
 * `outFiltered`: A choice whether to write an output file containing the filtered out reads by the previously cited filtering parameters
 
 * `cores`: The number of CPU cores to be used for a parallel processing
 
It is possible to apply the above-mentioned filtering functions subsequently to the data loading step in case users wish to explore the datasets before any data manipulation.


```{r data loading, echo=T, eval=F}
RepSeqData <- readAIRRSet(fileList = l,
                          fileFormat = "MiAIRR",
                          chain = "TRA",
                          sampleinfo = metadata,
                          keep.ambiguous = FALSE,
                          keep.unproductive = FALSE,
                          filter.singletons = FALSE,
                          aa.th = 8,
                          outFiltered = FALSE,
                          raretab = FALSE,
                          cores = 1L)

```


## The RepSeqExperiment object

### Architecture

Data loading with either one of `readFormatSet()` or `readAIRRSet()` allows the generation of a RepSeqExperiment object. This object is used as input in all of the functions.

It is composed of 4 slots, each containing a different type of information.

+ **assayData**: a data.table composed of all the alignment datasets and containing the following columns:

    - **sample_id**: Sample names

  - **V**: Variable gene name

  - **J**: Joining gene name

  - **VJ**: V-J gene combination using V and J gene names.
  
  - **ntCDR3**: Nucleotide CDR3 sequence

  - **aaCDR3**: Amino acid CDR3 sequence

  - **aaClone**: The full sequence including the V gene, the amino acid CDR3 sequence and the J gene

  - **ntClone**: The full sequence including the V gene, the nucleotide CDR3 sequence and the J gene 

  - **count**: The occurrence of each clone

This slot can be extracted using the `assay()` function.

```{r assay,  echo = T, eval=F}
assay(RepSeqData)
```

```{r view assay,  echo = F}
reactable::reactable(head(assay(RepSeqData)),  bordered = TRUE,  fullWidth = TRUE,pagination = FALSE,height = 300,
          defaultColDef = reactable::colDef(vAlign = "center",  align = "center", 
                                            minWidth = 200,
                                 headerStyle = list(background = "#f7f7f8")),
          columns = list(
            V= reactable::colDef(minWidth = 100),
            J=reactable::colDef(minWidth = 100),
            VJ=reactable::colDef(minWidth = 100),
            count=reactable::colDef(minWidth = 100)
          ))
```

+ **metaData**: a data frame containing the metadata information provided for the building of the RepSeqExperiment object, followed by a number of statistics calculated for each sample.

  These statistics include:
  
    * **nSequences**: the total number of sequences in a sample

  * **V**: The total number of V genes expressed in each sample
  
  * **J**: The total number of J genes

  * **VJ**: The total number of V-J gene combinations
  
  * **ntCDR3**: The number of unique nucleotide CDR3s

  * **aaCDR3**: The number of unique amino acid CDR3s
  
  * **aaClone**: The number of unique aaClones

  * **ntClone**: The number of unique ntClones
  
  * **Chao1**: Estimates undetected species using the information on the rarest species (the numbers of singletons and doubletons) (Chao, 1984).

  * **Improved Chao1**: An extension of **Chao1** which uses additional information, namely, the numbers of tripletons and quadrupletons  (Chiu et al., 2014).


This slot can be extracted using the `mData()` function.

```{r sdata,  echo = T, eval=F}
mData(RepSeqData)
```

```{r view metaData, echo = F}
tbl<- mData(RepSeqData)
tbl$iChao<- round(tbl$iChao, 2)
tbl$chao1<- round(tbl$chao1, 2)

reactable::reactable(tbl,  bordered = TRUE,  fullWidth = TRUE, 
          defaultColDef = reactable::colDef(vAlign = "center",  align = "center", 
                                            minWidth = 200,maxWidth = 110,
                                 headerStyle = list(background = "#f7f7f8"),
                                format = reactable::colFormat( separators = TRUE)),
     columns=list(
       sample_id= reactable::colDef(maxWidth = 150),
       cell_susbet= reactable::colDef(maxWidth = 150))
          )
 
```

- **otherData**: a list encompassing the output of a defined set of functions.

This slot can be extracted using the `oData()` function.

For instance, if the `outFiltered` parameter was set to TRUE during the RepSeqExperiment generation, filtered sequences would have been stored in this slot and can be viewed as followed.  

```{r odata,  echo = T, eval=F}
oData(RepSeqData)$filtered
```

In addition, `oData` contains a list of colors attributed for each group in the metadata file
```{r col,  echo = T, eval=T}
oData(RepSeqData)$label_colors
```



- **History**: a data frame registering all operations performed on the RepSeqExperiment object, such as the filtering and the down-sampling.

This slot can be extracted using the `History()` function.

```{r history,  echo = T, eval=F}
History(RepSeqData)
```

```{r view history, echo = F}
History(RepSeqData)[[1]][[1]] <- "data directory=./extdata/MiAIRR"
reactable::reactable(History(RepSeqData),
                     bordered = TRUE,  fullWidth = TRUE, 
          defaultColDef = reactable::colDef(
                                 headerStyle = list(background = "#f7f7f8")))
```

### RepSeqExp merging

`mergeRepSeq()` allows to combine two RepSeqExperiment objects encompassing different datasets. As such, a same **sample_id** cannot be in both datasets and should be renamed before using this function.

```{r}
l <- list.files(system.file(file.path('extdata/MiAIRR'),
                     package = 'AnalyzAIRR'),
                     full.names = TRUE)

dataset1 <- readAIRRSet(fileList = l[c(1:3)],
                       cores = 1L,
                       fileFormat = "MiAIRR",
                       chain = "TRA",
                       sampleinfo = NULL,
                       filter.singletons = FALSE,
                       outFiltered = FALSE)

dataset2 <- readAIRRSet(fileList = l[c(4:8)],
                       cores = 1L,
                       fileFormat = "MiAIRR",
                       chain = "TRA",
                       sampleinfo = NULL,
                       filter.singletons = FALSE,
                       outFiltered = FALSE)

dataset <- mergeRepSeq(a = dataset1, b = dataset2)
```
```{r ,  echo = F, eval=T}
tbl<- mData(dataset)
tbl$iChao<- round(tbl$iChao, 2)
tbl$chao1<- round(tbl$chao1, 2)

reactable::reactable(tbl,  bordered = TRUE,  fullWidth = TRUE, 
          defaultColDef = reactable::colDef(vAlign = "center",  align = "center", 
                                            minWidth = 200,maxWidth = 110,
                                 headerStyle = list(background = "#f7f7f8"),
                                format = reactable::colFormat( separators = TRUE)),
     columns=list(
       sample_id= reactable::colDef(maxWidth = 150),
       cell_suset= reactable::colDef(maxWidth = 150))
          )
```

