---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
'%>%' <-  kableExtra::'%>%'

```

# Devtools
The latest release of **AnalyzAIRR** can be installed from Github using **devtools**:
```{r github1, echo = TRUE, eval = FALSE}
install.packages("devtools")  
devtools::install_github("vanessajmh/AnalyzAIRR")
```

Load the **AnalyzAIRR** package
```{r load library, echo=T, eval=FALSE, warning=F, message=F}
library(AnalyzAIRR)
```

# Docker
A Docker image of the AnalyzAIRR tool encapsulating its dependencies, and required run environments is available on docker hub at vanessajmh/analyzairr.

To install the latest version, the following command line can be used:
```{bash, eval=F}
docker pull vanessajmh/analyzairr:latest
```

Images are versioned using tags. To install a specific version,  the following command line can be used:

```{bash, eval=F}
docker pull vanessajmh/analyzairr:v1.0.0

```



