---
title: "AnalyzAIRR"
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
'%>%' <-  kableExtra::'%>%'

```
# Introduction

**AnalyzAIRR** is a MiAIRR compliant R package developed to analyze bulk Ig/TCR repertoire datasets.

* It can be used by both biologists and bioinformaticians.

* It proposes an analytical routine starting with data exploration, leading to in-depth statistical comparisons with the aim of answering defined biological questions.

* It allows the calculation of a set of diversity measures and statistical metrics applicable at any level of granularity.

* It offers different types of data visualization and ready-to-publish graphics that can be easily personalized. 

This guide provides an overview of the pipeline proposed in **AnalyzAIRR**.

# Installation
## Devtools
The latest release of **AnalyzAIRR** can be installed from Github using **devtools**:
```{r github1, echo = TRUE, eval = FALSE}
install.packages("devtools")  
devtools::install_github("vanessajmh/AnalyzAIRR")
```

Load the **AnalyzAIRR** package
```{r load library, echo=T, eval=FALSE, warning=F, message=F}
library(AnalyzAIRR)
```

## Docker
A Docker image of the AnalyzAIRR tool encapsulating its dependencies, and required run environments is available on docker hub at vanessajmh/analyzairr.

To install the latest version, the following command line can be used:
```{bash, eval=F}
docker pull vanessajmh/analyzairr:latest
```

Images are versioned using tags. To install a specific version,  the following command line can be used:

```{bash, eval=F}
docker pull vanessajmh/analyzairr:v1.0.0

```


# Quick start

Details on supported data formats and the data loading process can be found [here](https://vanessajmh.github.io/AnalyzAIRR.github.io)

# Shiny interface

A Shiny web application was developed for **AnalyzAIRR** making it user-friendly for biologists with little or no background in bioinformatics.
The application can be either download from the [Github repository](https://github.com/vanessajmh/Shiny-AnalyzAIRR.git) or used directly at this [link](https://analyzairr.shinyapps.io/shiny-analyzairr/)



